Comparing vector-based and Bayesian memory models using large-scale datasets:\protect\\User-generated hashtag and tag prediction on Twitter and StackOverflow

### 2016-02-15 writeups/psycMethods/psycMethods.tex
foo

\begin{table}[!ht]
  \caption{ACT-R declarative memory model}
  \label{tabACTRModel}
  {\tabulinesep=1.2mm
    \begin{tabu}{ll}
      \hline
      Common Name &  Equation \\
      \hline
      Activation &	 	$A_{i} = B_{i} + \sum_{j \in c}^{} W_{j} S_{ji}$ \\
      Attentional Weight &	$W_{j} = \frac{W}{n}$ \\
      Base Level & 		$B_{i} = log \sum_{j=1}^{n} {t_{j}}^{-d}$ \\
      Constant Base Level &	$B_{i} = \frac{p_{i}}{1-p_{i}}$ \\
      Strength of Association &	$S_{ji} = log \frac{p(i|j)}{p(i|\overline{j})} \approx log \frac{p(i|j)}{p(i)} = log \frac{NN_{ji}}{N_{Row}(j)N_{Col}(i)}$ \\
      Recall Probability &	$P_{i} = \left( 1 + e^{\frac{\tau - A_{i}}{s}} \right )^{-1}$ \\
      \hline
    \end{tabu}
  }
\end{table}

@@

The resonance and decoding processes are inversions of each other.
Usually both resonance and decoding will return similar rank orderings of most likely chunks given context.
\textcite{Jones2007} recommends averaging the cosine values returned from each process for each chunk, and then rank ordering those averages to determine the chunk to retrieve.

@@

The fan effect describes a behavior of human memory where memory items are more difficult to retrieve if they have occurred in a larger number of different contexts.
\textcite{Rutledge2008} used \textcite{Anderson1974}'s classic experimental paradigm to test the fan effect.
In the task, participants are exposed to a set of person and place pairs (e.g., ``the hippie is in the park'').
They are later asked to recall if a particular person and place pair was presented to them before.
The main manipulation is the ``fan'' of the person and place for each pair.
The fan is defined as the number of different contexts that each item in the pair is presented in.
For example, if items were presented to the participant where the ``hippie'' was both in the ``park'' and in the ``store,'' then ``hippie'' would have a fan of two.
Recall accuracy decrease and time to recall increase as the sum total fan of the person and place increases.

Both the ACT-R declarative memory theory and vector-based models produce the fan effect due to the effect that fan has on the co-occurrence matrix.
As a particular word (e.g., ``hippie'') is seen in more and more contexts, the co-occurrence count of the word across all contexts is spread across multiple contexts.
This means that when a retrieval for a context given the word occurs, multiple contexts have relatively high activation.
This makes it more difficult to test if a signal (e.g., if a word appears in a specific context) is higher than the background noise (e.g., all other contexts that the word has appeared in),
which makes the retrieval more uncertain.

Both vector-based and Bayesian models produce the fan effect since they both are based on co-occurrences.
However, vector-based models do further processing on that matrix, reduce its dimensionality, and use a correlation instead of a log odds value to compute the activation.
Nonetheless, vector-based models still produce the fan effect because at their core they are still computing activations based on co-occurrences.

@@

This process was performed across all users in each of the 12 dataset slices for both the Twitter and StackOverflow popular-users datasets.
It was repeated for each of the decay rate values and the three different forms of $B_{i}$.
For the hybrid form, three different values of k were explored (1, 5, 10).

@@

Finally, model accuracy was analyzed across all dataset slices for StackOverflow and Twitter.
This was done by examining model accuracy at a specific decay rate value for each model.
Accuracy was assessed at each model's previously-determined best-fit values (\num{.43} and \num{.80} and \num{.80} for the optimized learning, standard form, and hybrid form) and additionally the standard form's default value (\num{0.5}).
Since decay rate values were explored at \num{0.1} increments between the 0 and 1 range, the decay rate values used to examine model accuracy were rounded to the nearest decay rate value that was included in the run.
Consequently, a best-fit value of \num{.4} was used for the optimized learning model instead of the \num{.43} average found from the dataset slices.
Aggregate accuracy was computed by taking the mean of every user's model accuracy across all users in all dataset slices.

@@

Accuracy for the hybrid form for the largest $k$ ($k=10$) is slightly higher than the smallest $k$ ($k=1$) (\numNoZero{.334} vs. \numNoZero{.329} \myCI{.0026}{.0023}{.0029}{3608}).
However accuracy cannot be distinguished between the hybrid form with the largest $k$ and the standard form (\numNoZero{.334} vs. \numNoZero{.334} \myCI{.0000}{-.0000}{.0001}{3609}).
This indicates that for the datasets analyzed in this study that the hybrid form with $k=10$ is a suitable replacement to the standard form (produces the same results) and uses less computational resources than the standard form.

@@

However the hybrid form with a reasonable $k$ is just as accurate as the standard form.
This is due to the fact that the majority of change in activation for a chunk occurs in the short period of time after the presentation of the chunk (a property of a power law distribution).
Once a chunk's presentations decay to the flat area of the power law distribution, not much change in activation for these presentations occurs, so they can be averaged and their presentation time can be discarded without much loss in fidelity.
Therefore if computational speed is a concern, one may wish to leverage the hybrid form at instead of the standard form.
Although qualitatively the standard form produces the cleanest and most natural effect of decay rate on accuracy, as shown in Figure \ref{figPriorSOQSliceDsStd}.
It also does not have a bifurcation at $k=1$.
So if one is a purest and computational speed is less of a concern, one may wish to use the standard form over the hybrid form.

@@

A simplified version of \textcite{Stanley2013}'s method was used for the results for this research.
Instead of subtracting the mean total activation for the top 5 tags for each post, the process is done in isolation for each model component.
So for the prior component, the activation value used in the logistic regression is the prior activation with the mean of the top 5 prior activations subtracted from this value.
The same process is used for the context component.
Decoupling the offset term and computing an offset for each model component in isolation means that an iterative approach is no longer necessary,
since the optimal weights of the terms are no longer needed when computing the offset.

@@

The transition point where accuracy for the Bayesian model outperformed the random permutation model was much higher for StackOverflow than Twitter.
This is most likely because StackOverflow has almost \num{100} times more unique tags than the Twitter popular-hashtags dataset.
The number of words in a StackOverflow post is also much higher than Twitter, which should have lowered the transition point.
However, the effect of the number of words in a post must be less than the effect of number of unique tags, as the transition point was still much higher for StackOverflow given that it should have lowered due to post size.

The compressed random permutation model should perform better with smaller corpus sizes, less words per document, more unique tags, and fewer number of tags per post.
However, since there are so many factors that determine whether the Bayesian or random permutation model will perform better,
and only two different datasets were examined in this research,
it is difficult to make any specific recommendations as far as exact transition cut points (e.g., corpus size, words per document) where either the Bayesian or random permutation model should be used.
Also, it is likely that the specific domain being studied interacts with the factors already mentioned to determine which model is better for that domain.
Nevertheless, if a researcher is working in a domain with limited data (e.g., small corpus size, large tag space) that produces a sparse co-occurrence matrix,
then it may be worthwhile to experiment with using a compressed random permutation model instead of the full Bayesian model.
Also newer models of $S_{ji}$ have been introduced that have a more local impact when a new chunk is observed (\parencite{thomson2013constraining}).
It would also be interesting to see how these local models of $S_{ji}$ perform while varying the number of observed associations.

>>>>

\begin{table}[!ht]
  \caption{ACT-R declarative memory model}
  \label{tabACTRModel}
  {\tabulinesep=1.2mm
    \begin{tabu}{ll}
      \hline
      Common Name &  Equation \\
      \hline
      Activation &	 	$A_{i} = B_{i} + \sum_{j \in c}^{} W_{j} S_{ji}$ \\
      Attentional Weight &	$W_{j} = \frac{W}{n}$ \\
      Base Level & 		$B_{i} = log \sum_{j=1}^{n} {t_{j}}^{-d}$ \\
      Strength of Association &	$S_{ji} = log \frac{p(i|j)}{p(i|\overline{j})} \approx log \frac{p(i|j)}{p(i)} = log \frac{NN_{ji}}{N_{Row}(j)N_{Col}(i)}$ \\
      Recall Probability &	$P_{i} = \left( 1 + e^{\frac{\tau - A_{i}}{s}} \right )^{-1}$ \\
      \hline
    \end{tabu}
  }
\end{table}

@@

The resonance and decoding processes are inversions of each other.
Usually both resonance and decoding will return similar rank orderings of most likely chunks given context.

@@

@@

This process was performed across all users in each of the 12 dataset slices for both the Twitter and StackOverflow popular-users datasets.
It was repeated for each of the decay rate values and the three different forms of $B_{i}$.
For the hybrid form, three different values of $k$ were explored (1, 5, 10).

@@

Finally, model accuracy was analyzed across all dataset slices for StackOverflow and Twitter.
This was done by examining model accuracy at a specific decay rate value for each model.
Accuracy was assessed at each model's previously-determined best-fit values (\num{.43}, \num{.80}, and \num{.80} for the optimized learning, standard form, and hybrid form) and additionally the standard form's default value (\num{0.5}).
Since decay rate values were explored at \num{0.1} increments between the 0 and 1 range, the decay rate values used to examine model accuracy were rounded to the nearest decay rate value that was included in the run.
Consequently, a best-fit value of \num{.4} was used for the optimized learning model instead of the \num{.43} average found from the dataset slices.
Aggregate accuracy was computed by taking the mean of every user's model accuracy across all users in all dataset slices.

@@

Accuracy for the hybrid form for the largest $k$ ($k=10$) is slightly higher than the smallest $k$ ($k=1$) (\numNoZero{.334} vs. \numNoZero{.329} \myCI{.0026}{.0023}{.0029}{3608}).
However accuracy cannot be distinguished between the hybrid form with the largest $k$ and the standard form (\numNoZero{.334} vs. \numNoZero{.334} \myCI{.0000}{-.0000}{.0001}{3609}).
This indicates that for the datasets analyzed in this study the hybrid form with $k=10$ is a suitable replacement to the standard form (produces the same results) and uses less computational resources than the standard form.

@@

However the hybrid form with a reasonable $k$ is just as accurate as the standard form.
This is due to the fact that the majority of change in activation for a chunk occurs in the short period of time after the presentation of the chunk (a property of a power law distribution).
Once a chunk's presentations decay to the flat area of the power law distribution, not much change in activation for these presentations occurs, so they can be averaged and their presentation time can be discarded without much loss in fidelity.
Therefore if computational speed is a concern, one may wish to leverage the hybrid form instead of the standard form.
Although qualitatively the standard form produces the cleanest and most natural effect of decay rate on accuracy, as shown in Figure \ref{figPriorSOQSliceDsStd}.
It also does not have a bifurcation at $k=1$.
So if one is a purest and computational speed is less of a concern, one may still wish to use the standard form over the hybrid form.

@@

A simplified version of \textcite{Stanley2013}'s method was used for this research.
Instead of subtracting the mean total activation for the top 5 tags for each post, the process is done in isolation for each model component.
So for the prior component, the activation value used in the logistic regression is the prior activation with the mean of the top 5 prior activations subtracted from this value.
The same process is used for the context component.
Decoupling the offset term and computing an offset for each model component in isolation means that an iterative approach is no longer necessary,
since the optimal weights of the terms are no longer needed when computing the offset.

@@

The compressed random permutation model should perform better with smaller corpus sizes, less words per document, more unique tags, and fewer number of tags per post.
However, since there are so many factors that determine whether the Bayesian or random permutation model will perform better,
and only two different datasets were examined in this research,
it is difficult to make any specific recommendations as far as exact transition cut points (e.g., corpus size, words per document) where either the Bayesian or random permutation model should be used.
Also, it is likely that the specific domain being studied interacts with the factors already mentioned to determine which model is better for that domain.
Nevertheless, if a researcher is working in a domain with limited data (e.g., small corpus size, large tag space) that produces a sparse co-occurrence matrix,
then it may be worthwhile to experiment with using a compressed random permutation model instead of the full Bayesian model.
Also newer models of $S_{ji}$ have been introduced that have a more local impact when a new chunk is observed \parencite{thomson2013constraining}.
It would also be interesting to see how these local models of $S_{ji}$ perform while varying the number of observed associations.
